---
title: "Shock detection in global fisheries production"
author: "Danielle Ferraro, UC Santa Barbara"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
bibliography: "brewfish.bib"
csl: "fish-and-fisheries.csl"
link_citations: true
---

# Introduction

The purpose of this document is to describe the methods for detecting shocks to fisheries production within the annual, country-level time series that has been collated for Track 1 within the Brewfish project. This code is largely rewritten from the GitHub repository associated with the Cottrell et al. [-@cottrell_food_2019] paper: "Food production shocks across land and sea."

Below is a first cut at rewriting the code. The original code was written to detect shocks across crop and livestock sectors in addition to fisheries and aquaculture sectors. [@cottrell_food_2019] used the Reg Watson marine capture database for fisheries landings in order to account for large-scale, small-scale, and IUU landings, and the FAO FishStat database for inland fisheries landings as well as inland and marine aquaculture production. The major modifications needed to use the code for our purposes will be:

- Use an updated aquaculture time series from FAO Global Production
- Use an updated fisheries time series from FAO Global Production
- Remove time series from other sectors, i.e. crops and livestock

# Setup

Install packages.
```{r packages, warning=FALSE, message=FALSE}

# Rich's
library(party)
library(randomForest)
library(splitstackshape)
library(plyr) # manipulation of data formatting
library(dplyr) # maniuplation of data formatting
library(sp) #spatial tool
library(spatial)
library(classInt)
#library(rworldmap) #if not using ggplot
library(RColorBrewer) #Color Brew Palettes
library(mapproj)
library(sf)#shapefile reading
library(ggalt)
library(ggplot2) #plotting
library(ggforce)
library(rgdal) 
library(ggmap)#creating ggmaps
library(devtools)
library(ggpubr) #for using the ggarrange function (displaying composite ggplots)
library(tidyr)#switching from wide to long format
library(zoo)# moving average functions 
library(countrycode) #adding countrycodes to data
library(colorspace) #adding colour palettes for plots
library(mgcv) #gams
library(car) # linear models
library(ggrepel) #ggrepel labels on point chart
# library(fANCOVA) #loess span fit function 
# library(forecast)
# library(earlywarnings)
# library(Kendall)
library(png)
library(grid) 
# library(readr)
# library(plot3D)
# library(circlize)
library(data.table)
library(purrr)

#Mine
library(here)
library(tidyverse)
```

Source scripts. 
```{r scripts}
walk(list.files(here("src"), pattern = "\\.R$", full.names = TRUE), source) # Sets file directories
```

# Import data

```{r import data}
country_key <- read_csv(file.path(dir_raw_data, "FAO", "production", "CL_FI_COUNTRY_GROUPS.csv")) 
production <- read_csv(file.path(dir_raw_data, "FAO", "production", "TS_FI_PRODUCTION.csv"), col_types = cols(UNIT = col_character())) %>% 
  rename_all(tolower)
```

# Tidy data

Append country names to production data. 

```{r}
production <- production %>% 
  left_join(country_key[,c("UN_Code", "Name_En")], by = c("country" = "UN_Code")) %>% 
  rename(country_name = Name_En)
```

Evaluate symbols. According to the FAO metadata (also available on NCEAS' Aurora server at `r file.path(dir_raw_data, "FAO", "production", "CL_FI_SYMBOL.csv")`, the different characters in the `SYMBOL` column correspond to the following notes:

Code|Symbol|Name|Description
-----|-----|-----|-----
1|.|UNKNOWN|Data not available; unobtainable; data not separately available but included in another category
2|0|NEGLIGIBLE|More than zero but less than half the unit used
3|-|ZERO|None; magnitude known to be nil or zero
4|F|ESTIMATED|FAO estimate from available sources of information or calculation based on specific assumptions
5|R|REPETITION|Repetition
6| |OFFICIAL|Official Data

```{r}
table(production$symbol) # Need to transform the negligibles (symbol=="0") and zeroes (symbol=="-")
production <- production %>% 
  mutate(quantity = case_when(symbol=="0" ~ 0.5,
                              symbol=="-" ~ 0,
                              TRUE ~ quantity))
```

Identify aquaculture vs. capture landings. According to the FAO metadata (also available on NCEAS' Aurora server at `r file.path(dir_raw_data, "FAO", "production", "CL_FI_PRODUCTION_SOURCE.csv")`, the different numbers in the `SOURCE` column correspond to the following types of production:

Identifier|Code|Name
-----|-----|-----
5|AQUACULTURE|Aquaculture production
1|FRESHWATER|Aquaculture production (freshwater)
2|BRACKISHWATER|Aquaculture production (brackishwater)
3|MARINE|Aquaculture production (marine)
4|CAPTURE|Capture production

```{r ID capture vs. aquaculture}
production <- production %>% 
  mutate(sector = if_else(source==4, "capture", "aquaculture"))
```

# Calculate aggregate production

Aggregate production per year, per country, per sector.
```{r calc aggregate production}
agg_production <- production %>% 
  group_by(country, country_name, year, sector) %>% 
  summarize(sum_quantity = sum(quantity)) %>% 
  ungroup(country, country_name, year, sector)
```

## Plot

Create a pdf with aggregated aquaculture and capture production by country.
```{r}
agg_production_list <- agg_production %>% 
  group_split(country_name)

agg_production_plots <- map(agg_production_list, 
                            ~ggplot(., aes(x = year, y = sum_quantity, color = sector)) + 
                              geom_hline(yintercept = 0, color = "darkgray", linetype = "dashed") +
                              geom_line() +
                              geom_point() +
                              scale_color_manual(values = c("darkred", "darkblue"), 
                                                 labels = c("Aquaculture", "Capture")) +
                              xlim(1950,2017) +
                              ylim(0,NA) +
                              labs(title = paste(.$country_name,"Fisheries Production"), 
                                   x = "Year", 
                                   y = "Quantity (tonnes)",
                                   color = NULL) +
                              ggpubr::theme_pubr() +
                              theme(legend.text = element_text(size=11))
)
agg_production_plots[[111]]
pdf(here("output", "fig_agg_production_bycountry.pdf"), paper = "a4")
agg_production_plots
dev.off()

ggsave(here("output", "fig_agg_production_bycountry.pdf"), gridExtra::marrangeGrob(grobs = agg_production_plots, widths nrow=2, ncol=1))


# Still wanna do: plot 2 figs per page
```

# Sensitivity Analysis - Cook's Distance

# References



